---
title: "Project report"
author: "Max Narvaez & Ryan Harvey"
date: "9/24/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

recognition.data <- read_csv("time_series_90s_d3_13_15_averaged_july_22.csv")
rec <- read_csv("time_series_90s_july13.csv")
```

## Instructions

The following document is a template of the report that will be due on Oct 1st at 9am, please change the information in lines 1-4 to have:

* Your project title
* The members of your group

The report *should* be uploaded as a PDF or an HTML document through Moodle. If you submit your project as an .Rmd you will be deducted *20 points* from your project grade.

Please delete the section *Instructions* from your final project

## Article review

In one or two paragraphs summarize the findings of the article that you will be reviewing/reproducing in your own words.

## Description of your dataset

Using a markdown table describe the variables from your dataset that you used for your analysis and whether they were in the original table or were obtained by transforming the original dataset. For variables that were not in the original dataset, briefly describe how you obtain them.

## Main conclusions 

Use well-written paragraphs and professional-looking tables and/or graphs to reproduce the main results of your article of interest. You should include at least the following points from each paper:

* *90's Music Cannon*:

    * The percent of songs from the 90's recognized currently decays with time.
```{r}
ggplot(recognition.data) +
    geom_line(aes(generation, recognition, color = artist_song)) +
    theme(legend.position = "none")

rec %>%
    filter(`-13` >= .9) %>%
    pivot_longer(c(`-13`, `-12`, `-11`, `-10`, `-9`, `-8`, `-7`, `-6`, `-5`, `-4`, `-3`, `-2`, `-1`, `0`, `1`, `2`, `3`, `4`, `5`, `6`, `7`, `8`, `9`, `10`),
                 names_to = "generation", values_to = "recognition") %>%
    transmute(song = artist_song, generation = generation, recognition = recognition) %>%
    filter(recognition > 0) %>%
    ggplot() +
    geom_line(aes(x = as.numeric(generation), recognition, color = song)) +
    scale_x_continuous(breaks = seq(-13, 10, by = 1)) +
    theme(legend.position = "none") +
    labs(x = "Age in year that song was released",
         y = "% of people who know song")
```
    * A graph of the top-5 /bottom-5 songs as they are currently remembered
    * A graph of the top-5 songs that were popular in the 90's but are rarely heard after that
    * A graph/table describing the change in recognition across different generations.
    
## Original question

Describe the question that you asked from the dataset and a visualization that describes the answer to such question. How do you interpret the dataset and how does it help answer your question.

    
    